---
title: "Data Analytics Project - Proposal"
author: "STAT 420, Summer 2023 - Soumya Nanda, Noam Isachar, Jonas Jansen"
date: ''
output:
  html_document:
  theme: readable
  toc: yes
  pdf_document: default
urlcolor: cyan
---

------------------------------------------------------------------------

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")
```

### 1. Team Members:

- Soumya Nanda - srnanda2
- Noam Isachar - noami2
- Jonas Jansen - jonasj2


### 2. Project Title:

Title Placeholder


### 3. Data File Description:

A brief statement of the business, science, research, or personal interest you have in the data set which you hope to explore

### 4. Data Background Information

Background information on the data sets, including specific citation of their source (so that I can also access it)


### 5. Team Interest

A brief statement of the business, science, research, or personal interest you have in the data set which you hope to explore

### 6. Data in R evidence

Evidence that the data can be loaded into R. Load the data, and print the first few values of the response variable as evidence.

```{r}
data <- read.csv("data/train.csv")

head(data)
```

