---
title: "Data Analytics Project - Proposal"
author: "STAT 420, Summer 2023 - Soumya Nanda, Noam Isachar, Jonas Jansen"
date: ''
output:
  pdf_document: default
  theme: readable
  toc: yes
  html_document:
urlcolor: cyan
---

------------------------------------------------------------------------

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.alin = "center")
```

### 1. Team Members:

-   Soumya Nanda - srnanda2
-   Noam Isachar - noami2
-   Jonas Jansen - jonasj2

### 2. Project Title:

#### Explaining and Predicting Song Popularity

### 3. Data File Description:

The dataset is called **Spotify_1Million_Tracks** and it contains 1.47M observations (songs) and 20 variables. We will try predicting the song's popularity which is determined by several factors, including the number of streams, user engagement, and how recently the song was released. The dataset also includes other details about each songs like the song's name and artist, and a few metrics from Spotify such as the danceability, energy and acousticness (numerical) as well as the key, time signature, year and genre (categorical).

Note that we chose this dataset and started to work on it before the staff sent the email that disallows response in the range between 0 and 100. We are excited about this dataset and analysis and don't think that the response closed range will take anything from the assignment or the learning experience so we hope to be able to go on with this dataset. If necessary, we can transform the response variable to the continuous range, or use the number of streams and plays from another dataset which is not limited in range but just positive.

### 4. Data Background Information

The dataset's source is here: <https://www.kaggle.com/datasets/amitanshjoshi/spotify-1million-tracks>[.\
](#0)The data was extracted from the Spotify platform using the Python library "Spotipy", which allows users to access music data provided via APIs. It includes over 1 million tracks with 19 features between 2000 and 2023. Also, there is a total of 61,445 unique artists and 82 genres in the data.

### 5. Team Interest

The questions we would like to answer are which features of a song impact its potential to be popular, and as a consequence is it possible to say how to write a song with the goal of it becoming a hit. Also we are interested at the challenge of predicting a song popularity reliably, potentially even prior to it's released, based on the song's musical characteristics and technical features.

### 6. Data in R evidence

```{r}
data <- read.csv("../data/spotify_data.csv")

head(data)
```
